**fellowship first markdown**

```{r}
library(tidyverse)
```


```{r}
# match data from Jeff Sackmann github

atp_matches2017 <- read_csv("data/atp_matches_2017.csv")
```



```{r}
# create 1st serve percent column

atp_matches2017 <- 
  atp_matches2017 %>%
  mutate(`first_serve_p_w` = w_1stIn/w_svpt) %>%
  mutate(`first_serve_p_l` = l_1stIn/l_svpt)
```


```{r}
# filter out non-GS matches

gs_matches2017 <-
  atp_matches2017 %>%
  filter(tourney_name == "Australian Open"| tourney_name == "Wimbledon"| tourney_name == "Roland Garros"| tourney_name == "US Open")  %>%
  within(., rm(first_serve_p_w, first_serve_p_l, `first_serve p_w`, `first_serve p_l`)) 

# rm( ) is for temporary variables, couldnt think of good names
```


```{r}
# attempt to pivot longer

gs_matches_long <-
  gs_matches2017 %>%
  pivot_longer(cols = c(w_first_serve_p, l_first_serve_p),
               winner_or_loser = "winner_or_loser",
               values_to = "serve_percentage")
```
*is there anyway to pivot longer without duplicating matches?*
*first shows winner's first serve percent, duplicate shows loser's first serve percent*




```{r}
# column for indicaor

gs_matches_long <-
  gs_matches_long %>%
  mutate(win = ifelse(winner_or_loser == "w_first_serve_p", 1, 0)) %>%
  within(., rm(winner_or_loser))
```


```{r}
# logistic for all GS matches -- trend is what I would've expected

ggplot(gs_matches_long, aes(x = serve_percentage, y = win)) +
  geom_jitter(height = 0.12, alpha = 0.1) +
  stat_smooth(method = "glm", method.args = c("binomial")) +
  stat_smooth()
```



```{r}
# example of the kind of thing the final app will do
# first subset the data based on players selected then plot logistic regression

players <-
  gs_matches_long %>%
  subset(winner_name == "Roger Federer"|
         winner_name == "Rafael Nadal")
  
ggplot(players, aes(x = serve_percentage, y = win, color = winner_name)) +
  geom_jitter(height = 0.12, alpha = 0.1) +
  stat_smooth(method = "glm", method.args = c("binomial")) +
  coord_cartesian(ylim = c(0, 1))

ggplot(gs_matches_long, aes(serve_percentage, win)) +
  stat_smooth(method = "glm", method.args = c("binomial")) +
  coord_cartesian(ylim = c(0, 1))

# is it possible to overlay these plots on each other?
```
*general trend for each is as expected*
*serve seems to be more of a deciding factor for Rafa*
*maybe since Fed's serve is a consistent weapon for him other factors play larger roles*


```{r}
allslams2017_long <-
  allslams2017 %>%
    pivot_longer(cols = c(player1, player2),
                names_to = "player_number",
                values_to = "player_name")

allslams2017_long %>%
  mutate(point_ind = case_when(
    player_number == "player1" & PointWinner == 1 ~ 1,
    player_number == "player1" & PointWinner == 2 ~ 0,
    player_number == "player2" & PointWinner == 1 ~ 0,
    player_number == "player2" & PointWinner == 2 ~ 1))

roger_rafa <-
  allslams2017_long %>%
  subset(., player_name == "Rafael Nadal"| player_name == "Roger Federer")


```
*same potential problem as before - duplicates each point*
*this gives us a player variable but im not sure if this will give us problems later*

*how can we merge importance with allslams2017_long?*

 
```{r}
# serverwin vs importance for Roger, Rafa
# same strategy as before -- subset atp_df with selected players then plot

#big 3
selected_players <-
  atp_df %>%
  pivot_longer(., cols = c(player1, player2),
               names_to = "player_num",
               values_to = "player_name") %>%
  subset(player_name == "Roger Federer" |
         player_name == "Rafael Nadal" |
      player_name == "Novak Djokovic") %>%
  select(player_name, player_num, PointServer, serverwin, everything()) %>%
  mutate(returnerwin = case_when(
    PointServer == 1 & PointWinner == 1 ~ 0,
    PointServer == 1 & PointWinner == 2 ~ 1,
    PointServer == 2 & PointWinner == 1 ~ 1,
    PointServer == 2 & PointWinner == 2 ~ 0)) %>%
  filter((player_num == "player1" & PointServer == 1) |
      player_num == "player2" & PointServer == 2)

# all players
all_players <- 
  atp_df %>%
  pivot_longer(., cols = c(player1, player2),
               names_to = "player_num",
               values_to = "player_name") %>%
  select(player_name, player_num, PointServer, serverwin, everything()) %>%
  mutate(returnerwin = case_when(
    PointServer == 1 & PointWinner == 1 ~ 0,
    PointServer == 1 & PointWinner == 2 ~ 1,
    PointServer == 2 & PointWinner == 1 ~ 1,
    PointServer == 2 & PointWinner == 2 ~ 0)) %>%
  filter((player_num == "player1" & PointServer == 1) |
      player_num == "player2" & PointServer == 2)
  

## this is pretty surprising though....I would not have expected such
## a strong pattern anyway
## it might be interesting to look at all the matches for the year, not
## just GS matches and see if the pattern holds up.
ggplot(selected_players, aes(importance, serverwin, color = player_name)) +
  geom_jitter(height = 0.12, alpha = 0.25) +
  stat_smooth(method = "glm", method.args = c("binomial")) 


ggplot(selected_players, aes(Speed_MPH, serverwin, color = player_name)) +
  geom_jitter(height = 0.12, alpha = 0.25) +
  stat_smooth(method = "glm", method.args = c("binomial"))
# this one doesnt quite make sense to me
# the 0's in Speed_MPH are messing this up a bit

ggplot(selected_players, aes(importance, returnerwin, color = player_name)) +
  stat_smooth(method = "glm", method.args = c("binomial"))
# pretty interesting that each players' trend flips for returning

# no apparent pattern for big 3
ggplot(selected_players, aes(importance, Speed_MPH, color = player_name)) +
  geom_point()

#no apparent patten for all players
ggplot(all_players, aes(importance, Speed_MPH)) +
  geom_point()
```
*same surprising result*
*especially with Roger and Rafa I would expect better performance at higher importance*
*maybe the higher importance motivates the returner more than the server for some reason??*


```{r}
#replace 0's in Speed_MPH with NA
cleaned_selected_players <-
  selected_players %>%
  mutate(ServeSpeed = ifelse(Speed_MPH == 0 & doublefault == 0, NA, 
                            ifelse(Speed_MPH == 0 & doublefault == 1, 0,
                                   ifelse(Speed_MPH != 0, parse_number(Speed_MPH), NULL))))

cleaned_selected_players <-
  selected_players %>%
  mutate(ServeSpeed = case_when(
    Speed_MPH == 0 & doublefault == 0 ~ NA,
    Speed_MPH == 0 & doublefault == 1 ~ 0,
    Speed_MPH != 0 ~ Speed_MPH
  ))
# need help here, cant figure out a way to get rid of 0's if doublefault == 0 and Speed_MPH == 0

# No NA's in Speed_MPH
any(is.na(all_players$Speed_MPH))
```


```{r}
# does importance impact return depth?

# big 3 (make indicator for whether it was deep)
selected_players %>%
  mutate(ReturnDeep = ifelse(ReturnDepth == "D", 1, 0), na.rm = T) %>%
  ggplot(., aes(importance, ReturnDeep)) +
  stat_smooth(method = "glm", method.args = "binomial") +
  coord_cartesian(ylim = c(0, 1))

# all players (make indicator for whether it was deep)
all_players %>%
  mutate(ReturnDeep = ifelse(ReturnDepth == "D", 1, 0), na.rm = T) %>%
  ggplot(., aes(importance, ReturnDeep)) +
  stat_smooth(method = "glm", method.args = "binomial") +
  coord_cartesian(ylim = c(0, 1))
```
*looks like big 3 are slightly better at getting the return deep than all players*
